Advancing spatially explicit fisheries management with age-structured species distribution models

### 2025-01-11 paper.qmd
Update paper.qmd

Species distribution models (SDMs) are widely used in fisheries and ecology, providing insights into species' spatial distributions. While many SDMs incorporate external information, such as environmental or habitat features, the majority overlook age-specific information, despite the known influence of age structure on fish distribution and abundance. Including age information may be particularly valuable for species exhibiting age-related spatial patterns driven by ontogenetic shifts, recruitment dynamics, and selective fishing pressures. Here, we develop an age-structured SDM framework using data from the west coast of the USA, and concentrate on two species with distinct life histories: North Pacific hake (*Merluccius productus*) and sablefish (*Anoplopoma fimbria*). Our framework uses spatiotemporal models to predict age-specific distributions. We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlight that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes. Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2010 and 2014 cohorts; sablefish from 2013, 2016, and 2020 cohorts). Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk, helping increase the efficiency of fisheries. Our framework is applicable to a wide range of survey types and platforms around the world, and supports the development of spatially explicit management strategies and optimal allocation of fishing effort.

@@

Species distribution models (SDMs) are widely used in marine ecology to understand and predict the spatial distribution of species based on environmental and demographic variables [@melo-merino2020]. These models have informed management decisions for conservation and resource allocation, especially in the face of changing environmental conditions and human impacts, such as fishing pressure [@guisan2005; @elith2009]. Recent advances in SDM applications in the marine environment have incorporated external drivers such as temperature [@fredston2021], habitat [@phillips2017; @laman2018]. A second use of these models has been to predict temporal trends in species biomass from spatiotemporal fisheries independent survey data [@thorson2015]. Trends estimated from SDMs may be included in integrated population models (stock assessments) for commercially fished species, or used to identify risks to populations of conservation concern. Recent developments in these models have included depth [@johnson2019] and habitat variables [@cao2017] to improve trend estimation.

@@

Incorporating age structure into SDMs may provide more accurate predictions by capturing spatial and temporal variability related to life history traits, age-specific habitat preferences, and population dynamics. Fish often exhibit ontogenetic shifts, where younger individuals occupy different habitats than adults due to factors like predator avoidance, habitat preferences, and diet changes [@werner1984; @ciannelli2004]. These shifts can create age-specific distributions across habitats, particularly for species that undergo distinct life history phases. Additionally, recruitment pulses and migration patterns vary by age, influencing the spatial distribution of younger versus older fish. Fishing selectivity further accentuates these patterns by disproportionately removing older age classes, which alters the population structure and potentially shifts spatial distributions [@hilborn1992; @methot2013]. By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relationships.

@@

On the west coast of the USA, the West Coast Groundfish Bottom Trawl Survey (WCGBTS) has been an annual survey conduced from 2003 - present. The WCGBTS is designed to estimate the abundance, size, and age composition of groundfish species important to commercial and recreational fisheries found in near-bottom habitats on the west coast of the USA (Keller et al., 2017). The survey effort is concentrated in summer months and has been conducted annually since 2003 (here we use the data through 2018; data are publicly available at https://www.nwfsc.noaa.gov/data). Importantly, the random stratified sampling design, spatial and seasonal coverage, effort, and gears have remained relatively constant within the period we analyse. Like many other surveys around the world, no WCGBTS survey occured in 2020 because of the Covid-19 pandemic. Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish. \[more on biological gradients\] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years. Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada (Grandin et al. 2024) and sablefish is frequently assessed because of its high commercial value. For our analysis, we used hake collected 2007 -- 2019 (mean 650.7 individuals sampled per year) and sablefish collected 2003 -- 2023 (mean 1314.4 individuals sampled per year).

@@

As a first validation, we leveraged the natural recursive element of our data to quantify the ability of our models to predict the future distribution by age. For each of the species - age models constructed above, we made predictions to the survey locations in the following year (e.g. age 3 hake in years 2007 -- 2018 was used to predict the distribution of age 4 hake in 2008 -- 2019). We related the predictions $\hat{u}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{u}_{\boldsymbol{s},t}$. We quantified the relationship between predictions and observations by fitting a simple Poisson GLM, where counts of fish of age $a+1$ in year $t+1$ were treated as the response and $log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)$. The exponentiated slope parameter $exp(\beta_{1})$ represents a change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$, and can roughly be interpreted as an estimate of relative survival. As a diagnostic check and to understand whether there was potential temporal patterns in the response, we fit a second model as a GLMM with R package glmmTMB [@brooks2017]; this second model was identical to our Poisson GLM with random effects in $\beta_{0}$ and $\beta_{1}$, where year was used as the grouping variable.

@@

When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, there is a decrease in the estimated spatial and spatiotemporal variances as well as a decrease in the spatial range (@fig-spatial-parameters). Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, meaning that spatial fields estimated for older fish are less variable. Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ \~ 0.13). The decline in range with age implies that spatial distributions also become more concentrated as a function of age (@fig-spatial-parameters).

@@

```{r label="fig-glm-coefficients", fig.cap="Estimated coefficients relating predicted densities of age a fish in year t to observed numbers the following year. The coefficients are interpreted with a multiplicate effect, such that a coefficient of 0.5 translates to an effect of exp(0.5) = 65% increase on the expected count in the future.", echo=FALSE}
knitr::include_graphics("plots/glm_coefficients.png")
```

>>>>

Species distribution models (SDMs) are widely used in fisheries and ecology, providing insights into species' spatial distributions. While many SDMs incorporate external information, such as environmental or habitat features, the majority overlook age-specific information, despite the known influence of age structure on fish distribution and abundance. Including age information may be particularly valuable for species exhibiting age-related spatial patterns driven by ontogenetic shifts, recruitment dynamics, and selective fishing pressures. Here, we develop an age-structured SDM framework using data from the west coast of the USA, and concentrate on two species with distinct life histories: North Pacific hake (*Merluccius productus*) and sablefish (*Anoplopoma fimbria*). Our framework uses spatiotemporal models to predict age-specific distributions. We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlight that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes. Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2010 and 2014 cohorts; sablefish from 2013, 2016, and 2020 cohorts). Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk in space, helping increase the efficiency of fisheries. Our framework is applicable to a wide range of survey types and platforms around the world, and supports the development of spatially explicit management strategies and optimal allocation of fishing effort.

@@

Species distribution models (SDMs) are widely used in marine ecology to understand and predict the spatial distribution of species based on environmental and demographic variables [@melo-merino2020]. These models have informed management decisions for conservation and resource allocation, especially in the face of changing environmental conditions and human impacts, such as fishing pressure [@guisan2005; @elith2009]. Recent advances in SDM applications in the marine environment have incorporated external drivers such as temperature [@fredston2021] and habitat features [@phillips2017; @laman2018]. A second use of these models has been to predict temporal trends in species biomass from spatiotemporal fisheries independent survey data [@thorson2015]. Trends estimated from SDMs may be included in integrated population models (stock assessments) for commercially fished species, or used to identify risks to populations of conservation concern. Recent developments in these models have included depth [@johnson2019] and habitat variables [@cao2017] to improve trend estimation.

@@

Incorporating age structure into SDMs may provide more accurate predictions by capturing spatial and temporal variability related to life history traits, age-specific habitat preferences, and population dynamics. Fish often exhibit ontogenetic shifts, where younger individuals occupy different habitats than adults due to factors like predator avoidance, habitat preferences, and diet changes [@werner1984; @ciannelli2004]. These shifts can create age-specific distributions across habitats, particularly for species that undergo distinct life history phases. Additionally, recruitment pulses and migration patterns vary by age, influencing the spatial distribution of younger versus older fish. Fishing selectivity often further accentuates these patterns by disproportionately removing larger, older fish, which alters the population structure and potentially shifts spatial distributions [@hilborn1992; @methot2013]. By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relati

@@

On the west coast of the USA, the West Coast Groundfish Bottom Trawl Survey (WCGBTS) has been an annual survey conduced from 2003 - present. The WCGBTS is designed to estimate the abundance, size, and age composition of groundfish species important to commercial and recreational fisheries found in near-bottom habitats on the west coast of the USA (Keller et al., 2017). The survey effort is concentrated in summer months and has been conducted annually since 2003 (here we use the data through 2018; data are publicly available at https://www.nwfsc.noaa.gov/data). Importantly, the random stratified sampling design, spatial and seasonal coverage, effort, and gears have remained relatively constant within the period we analyze. Like many other surveys around the world, no WCGBTS survey occurred in 2020 because of the Covid-19 pandemic. Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish. \[more on biological gradients\] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years. Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada (Grandin et al. 2024) and sablefish is frequently assessed because of its high commercial value. For our analysis, we used hake collected 2007 -- 2019 (mean 650.7 individuals sampled per year) and sablefish collected 2003 -- 2023 (mean 1314.4 individuals sampled per year).

@@

As a first validation, we leveraged the natural recursive element of our data to quantify the ability of our models to predict the future distribution by age. For each of the species - age models constructed above, we made predictions to the survey locations in the following year (e.g. age 3 hake in years 2007 -- 2018 was used to predict the distribution of age 4 hake in 2008 -- 2019). We related the predictions $\hat{u}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{u}_{\boldsymbol{s},t}$. We quantified the relationship between predictions and observations by fitting a simple Poisson GLM, where counts of fish of age $a+1$ in year $t+1$ were treated as the response and $log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)$. The exponentiated slope parameter $exp(\beta_{1})$ represents a change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$. As a diagnostic check and to understand whether there was potential temporal patterns in the response, we fit a second model as a GLMM with R package glmmTMB [@brooks2017]; this second model was identical to our Poisson GLM with random effects in $\beta_{0}$ and $\beta_{1}$, where year was used as the grouping variable.

@@

When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, there is a decrease in the estimated spatial and spatiotemporal variances as well as a decrease in the spatial range with increasing age (@fig-spatial-parameters). Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, meaning that spatial fields estimated for older fish are less variable. Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ \~ 0.13). The decline in range with age implies that spatial distributions also become more concentrated as a function of age (@fig-spatial-parameters).

@@

```{r label="fig-glm-coefficients", fig.cap="Estimated coefficients relating predicted densities of age a fish in year t to observed numbers the following year. The coefficients and 95% confidence intervals (vertical lines) are estimated in log space and presented here in normal space.", echo=FALSE}
knitr::include_graphics("plots/glm_coefficients.png")
```
